---
title: "Parking Tickets Analysis in Toronto, ON"
output:
  html_document:
    df_print: paged
---

# 2.0 Datasets

The Parking Tickets dataset from the City of Toronto's Open Data Portal (https://open.toronto.ca/dataset/parking-tickets/) was used for this analysis. 

Please note that parking ticket data from 2009 was not included in this analysis. In order to submit something at all, this data was omitted as I was unable to import the 2009 Parking Tags CSV file into R.

```{r load, message=FALSE, warning=FALSE}
#Load required data from the RData file
load(file = "Kate_Woods_ID13025_data.RData")

#Load required packages
library(leaflet)
library(dplyr)
library(plyr)
library(sf)
library(ggplot2)
library(GISTools)
library(ggsn)
library(rgdal)
library(tidyverse)
library(spatstat)
library(maptools)
library(spdep)
library(knitr)
library(gridExtra)
library(lubridate)
```


# 3.1 Analysis

### 3.1.1 Top 20 ticket infractions (frequency)

```{r}
library(plyr)

inf_freq <- plyr::count(park, 'infraction_description')

df <- inf_freq[order(-inf_freq$freq),]

top_20 <- head(df, n=20)
top_20
```

### 3.1.2 Top 20 ticket infractions (revenue)

### 3.1.3 Total revenue for all tickets

```{r}
sum(park$set_fine_amount)
```


### 3.1.4.a Distance to closest parking lots for top 20 infractions

### 3.1.4.b Distance to closest TTC stop for top 20 infractions

### 3.1.5.a Impact of day of week in all infractions

```{r}
park$day <- day(ymd(park$date_of_infraction))

park$wday <- wday(ymd(park$date_of_infraction), label=TRUE, abbr=FALSE)

park$month <- month(ymd(park$date_of_infraction), label=TRUE, abbr=FALSE)

park$year <- year(ymd(park$date_of_infraction))

wday_freq <- plyr::count(park, 'wday')
wday_freq
```
### 3.1.5.b Impact of month of week in all infractions

```{r}
month_freq <- plyr::count(park, 'month')
month_freq
```

### 3.1.5.c Impact of season of week in all infractions


# 3.3 Visualization

### 3.3.1.a.i Distribution of infractions by year

```{r}
year_freq <- plyr::count(park, 'year')

ggplot(year_freq, aes(x=year, y=freq )) +
  geom_bar(stat="identity")
```


### 3.3.1.a.ii Distribution of infractions by month

```{r}
ggplot(month_freq, aes(x=month, y=freq )) +
  geom_bar(stat="identity")
```


### 3.3.1.b Distribution of top 20 infractions by fines

### 3.3.2.a Geographic distribution (location) of top 20 infractions (count)

### 3.3.2.b Geographic distribution by ward for top 20 infractions (count)

```{r}
#Import Toronto wards shapefile
toronto <- st_read(dsn = "/Users/katewoods14/id13025wa/Ward Data/25-ward-model-december-2018-wgs84-latitude-longitude", 
                   layer = "WARD_WGS84")


plot(toronto)
```


### 3.3.2.c Geographic distribution by ward for top 20 infractions (revenue)

### 3.3.2.d Any relevant observations in relation to socio demographic profiles

